---
title: "Gender Dataset PCA analysis"
output: html_document
date: "2023-02-15"
editor_options: 
  chunk_output_type: console
---

```{r}
knitr::opts_chunk$set(echo = TRUE)

#Tanay Dangaich

#import libraries
library(readr)
library(tidyverse)
library(ggplot2)
library(ggcorrplot)

#read csv file and store in sparr
gender <- read_csv("~/Documents/Courses/Multivariate Analysis/gender.csv")
gender
#male matrix, colMeans, covariance and correlation


df <- as.data.frame(gender)
View(df)

df_1 = df
df_1 = df_1[, -9:-10]
df_1 = df_1[, -5:-7]
View(df_1)

male = filter(df_1, Gender == "male")
male.nums <- male[,-1]
View(male.nums)

#colMeans
colMeans(male.nums)
#covariance
cov(male.nums)
#correlation
cor_male <- cor(male.nums)
cor_male
ggcorrplot(cor_male)
```


>Inference:
>The height and weight are highly correlated for males.
>Income and age are correlated, the higher the age, the more they are earning.
>Age and weight also have a positive correlation.



```{r}
#female matrix, colMeans, covariance and correlation
female = filter(df_1, Gender == "female")
female.nums <- df_1[,-1]
View(female.nums)

#colMeans
colMeans(female.nums)
#covariance
cov_female <- cov(female.nums)
cov_female

#correlation
cor_female <- cor(female.nums)
cor_female
ggcorrplot(cor_female)
```



>Inference:
>The height and weight are highly correlated for females.
>Income and age are correlated, the higher the age, the more they are earning.
>Age and weight also have a positive correlation.

```{r}
knitr::opts_chunk$set(echo = TRUE)
library(readr)
library(ggplot2)
gender <- read_csv("~/Documents/Courses/Multivariate Analysis/gender.csv")
str(gender)


df <- as.data.frame(gender)
View(df)

df_1 = df
df_1 = df_1[, -9:-10]
df_1 = df_1[, -5:-7]
View(df_1)

library(tidyverse)

df_1 <- df_1 %>% 
  mutate(
    Gender = ifelse(Gender %in% c("male"), "M", "W") 
  )

View(df_1)

boxplot(df_1[,2:5])

stars(df_1,labels = df_1$Gender)


library(SciViews)
library(scatterplot3d)
library(car)
library(GGally)
library(lattice)
library(ggplot2)
library(ggridges)
library(ggvis)
library(ggthemes)
library(cowplot)
library(gapminder)
library(gganimate)
library(dplyr)
library(tidyverse)
library(grid)
library(gridExtra)
library(RColorBrewer)
library(hexbin)
names(df_1)

ggscatmat(df_1, columns=2:5, color="Gender")

# ggplot
#Using ggplot to plot different variables against survivorship of birds
ggplot(df_1, aes(x=Gender,y=Age)) + geom_point(aes(color=Gender))
ggplot(df_1, aes(x=Gender,y=`Height (cm)`)) + geom_point(aes(color=Gender))
ggplot(df_1, aes(x=Gender,y=`Weight (kg)`)) + geom_point(aes(color=Gender))
ggplot(df_1, aes(x=Gender,y=`Income (USD)`)) + geom_point(aes(color=Gender))



ggplot(df_1, aes(x=Age, y=`Income (USD)`)) + geom_point(aes(color=Gender))

ggplot(df_1, aes(x=Age,y=`Income (USD)`)) + geom_point(aes(color=Gender), pch=17) + 
  labs(x="Age", y="Income in USD", title="Gender data") 





# bar chart
ggplot(df_1, aes(Age)) + geom_bar(position="stack",aes(color=Gender)) 
ggplot(df_1, aes(`Height (cm)`)) + geom_bar(position="stack",aes(color=Gender))
ggplot(df_1, aes(`Weight (kg)`)) + geom_bar(position="stack",aes(color=Gender))
ggplot(df_1, aes(`Income (USD)`)) + geom_bar(position="stack",aes(color=Gender))

ggplot(df_1, aes(Age)) + facet_grid(.~Gender) + geom_bar(position="dodge",aes(color=Gender))
ggplot(df_1, aes(`Height (cm)`)) + facet_grid(.~Gender) + geom_bar(position="dodge",aes(color=Gender))
ggplot(df_1, aes(`Weight (kg)`)) + facet_grid(.~Gender) + geom_bar(position="dodge",aes(color=Gender))
ggplot(df_1, aes(`Income (USD)`)) + facet_grid(.~Gender) + geom_bar(position="dodge",aes(color=Gender))



# histogram
ggplot(df_1, aes(Age))+geom_histogram(aes(color=Gender))
ggplot(df_1, aes(`Height (cm)`))+geom_histogram(aes(fill = after_stat(count)))
ggplot(df_1, aes(`Weight (kg)`))+geom_histogram(aes(fill = after_stat(count)))




# violin plot 
ggplot(df_1, aes(x=Gender, y=Age)) + geom_violin(aes(color=Gender))
ggplot(df_1, aes(x=Gender, y=`Height (cm)`)) + geom_violin(aes(color=Gender))
ggplot(df_1, aes(x=Gender, y=`Weight (kg)`)) + geom_violin(aes(color=Gender))
ggplot(df_1, aes(x=Gender, y=`Income (USD)`)) + geom_violin(aes(color=Gender))



# box plot
ggplot(df_1, aes(x=Gender, y=Age)) + geom_boxplot(aes(color=Gender))
ggplot(df_1, aes(x=Gender, y=`Height (cm)`)) + geom_boxplot(aes(color=Gender))
ggplot(df_1, aes(x=Gender, y=`Weight (kg)`)) + geom_boxplot(aes(color=Gender))
ggplot(df_1, aes(x=Gender, y=`Income (USD)`)) + geom_boxplot(aes(color=Gender))



# density plot and ggridges
ggplot(df_1, aes(x=Age, fill=Gender, color=Gender)) + geom_density(alpha=0.3, aes(y=..scaled..)) 

ggplot(df_1, aes(x=Age, y=Gender)) + geom_density_ridges(aes(fill=Gender, color=Gender))



# hexbin
ggplot(df_1, aes(x=Age, y=Gender)) + geom_hex(aes(color=Gender)) 




# with ggthemes (see also ggsci, ggthemr)
lastplot <- ggplot(df_1, aes(x=Age,y=`Income (USD)`)) + geom_point(aes(color=Gender), pch=18) + 
  labs(x="Age", y="Income (USD)", title="Gender Data") 

lastplot + theme_bw()
lastplot + theme_cowplot()
lastplot + theme_dark()
lastplot + theme_economist()
lastplot + theme_fivethirtyeight()
lastplot + theme_tufte()
lastplot + theme_wsj()
```


>Inference:
>The height and weight are highly correlated for females.
>Income and age are correlated, the higher the age, the more they are earning.
>Age and weight also have a positive correlation.


```{r}
library(readr)
library(caret)

gender <- read_csv("~/Documents/Courses/Multivariate Analysis/gender.csv")
str(gender)
attach(gender)
View(gender)

df <- as.data.frame(gender)
View(df)

df_1 = df
df_1 = df_1[, -9:-10]
df_1 = df_1[, -5:-7]
View(df_1)

#dmy <- dummyVars(" ~ .", data = df_1)
#trsf <- data.frame(predict(dmy, newdata = df_1))
#View(trsf) 

#Get the Correlations between the measurements
cor(df_1[-1])
# Using prcomp to compute the principal components (eigenvalues and eigenvectors). With scale=TRUE, variable means are set to zero, and variances set to one
gender_pca <- prcomp(df_1[,-1],scale=TRUE)
gender_pca
summary(gender_pca)
# sample scores stored in sparrows_pca$x
# singular values (square roots of eigenvalues) stored in sparrow_pca$sdev
# loadings (eigenvectors) are stored in sparrows_pca$rotation
# variable means stored in sparrows_pca$center
# variable standard deviations stored in sparrows_pca$scale
# A table containing eigenvalues and %'s accounted, follows
# Eigenvalues are sdev^2
(eigen_gender <- gender_pca$sdev^2)
names(eigen_gender) <- paste("PC",1:4,sep="")
eigen_gender
sumlambdas <- sum(eigen_gender)
sumlambdas
propvar <- eigen_gender/sumlambdas
propvar
cumvar_gender <- cumsum(propvar)
cumvar_gender
matlambdas <- rbind(eigen_gender,propvar,cumvar_gender)
matlambdas

rownames(matlambdas) <- c("Eigenvalues","Prop. variance","Cum. prop. variance")
matlambdas
round(matlambdas,4)
summary(gender_pca)
gender_pca$rotation
print(gender_pca)
## Sample scores stored in gender_pca$x
gender_pca$x
# Identifying the scores by their gender
gendertyp_pca <- cbind(data.frame(Gender),gender_pca$x)
View(gendertyp_pca)
# Means of scores for all the PC's classified by Gender
tabmeansPC <- aggregate(gendertyp_pca[,2:5],by=list(Gender=df_1$Gender),mean)
View(tabmeansPC)
tabmeansPC <- tabmeansPC[rev(order(tabmeansPC$Gender)),]
View(tabmeansPC)
tabfmeans <- t(tabmeansPC[,-1])
tabfmeans
colnames(tabfmeans) <- t(as.vector(tabmeansPC[1]$Gender))
tabfmeans
# Standard deviations of scores for all the PC's classified by Gender
tabsdsPC <- aggregate(gendertyp_pca[,2:5],by=list(Gender=df_1$Gender),sd)
tabfsds <- t(tabsdsPC[,-1])
colnames(tabfsds) <- t(as.vector(tabsdsPC[1]$Gender))
tabfsds
t.test(PC1~df_1$Gender,data=gendertyp_pca)
t.test(PC2~df_1$Gender,data=gendertyp_pca)
t.test(PC3~df_1$Gender,data=gendertyp_pca)
t.test(PC4~df_1$Gender,data=gendertyp_pca)
## F ratio tests
var.test(PC1~df_1$Gender,data=gendertyp_pca)
var.test(PC2~df_1$Gender,data=gendertyp_pca)
var.test(PC3~df_1$Gender,data=gendertyp_pca)
var.test(PC4~df_1$Gender,data=gendertyp_pca)
# Levene's tests (one-sided)
library(car)
(LTPC1 <- leveneTest(PC1~df_1$Gender,data=gendertyp_pca))
(p_PC1_1sided <- LTPC1[[3]][1]/2)
(LTPC2 <- leveneTest(PC2~df_1$Gender,data=gendertyp_pca))
(p_PC2_1sided=LTPC2[[3]][1]/2)
(LTPC3 <- leveneTest(PC3~df_1$Gender,data=gendertyp_pca))
(p_PC3_1sided <- LTPC3[[3]][1]/2)
(LTPC4 <- leveneTest(PC4~df_1$Gender,data=gendertyp_pca))
(p_PC4_1sided <- LTPC4[[3]][1]/2)

# Plotting the scores for the first and second components
plot(gendertyp_pca$PC1, gendertyp_pca$PC2,pch=ifelse(gendertyp_pca$Gender == "M",1,16),xlab="PC1", ylab="PC2", main="131 people against values for PC1 & PC2")
abline(h=0)
abline(v=0)
legend("bottomleft", legend=c("Male","Female"), pch=c(1,16))
plot(eigen_gender, xlab = "Component number", ylab = "Component variance", type = "l", main = "Scree diagram")
plot(log(eigen_gender), xlab = "Component number",ylab = "log(Component variance)", type="l",main = "Log(eigenvalue) diagram")
print(summary(gender_pca))
diag(cov(gender_pca$x))
xlim <- range(gender_pca$x[,1])
xlim
gender_pca$x[,1]
gender_pca$x
plot(gender_pca$x,xlim=xlim,ylim=xlim)
gender_pca$rotation[,1]
gender_pca$rotation
plot(df_1[,-1])
gender_pca$x
plot(gender_pca)
#get the original value of the data based on PCA
center <- gender_pca$center
scale <- gender_pca$scale
new_gender <- as.matrix(df_1[,-1])
new_gender
drop(scale(new_gender,center=center, scale=scale)%*%gender_pca$rotation[,1])
predict(gender_pca)[,1]
#The aboved two gives us the same thing. predict is a good function to know.
df_1$Gender <- as.factor(df_1$Gender)
out <- sapply(1:4, function(i){plot(df_1$Gender,gender_pca$x[,i],xlab=paste("PC",i,sep=""),ylab="Gender")})
pairs(gender_pca$x[,1:4], ylim = c(-6,4),xlim = c(-6,4),panel=function(x,y,...){text(x,y,df_1$Gender)})

# Better Ways to Visualize

library(factoextra)
library(FactoMineR)
library(ggfortify)
library(psych)
library(corrplot)
library(devtools)

# Correlation
pairs.panels(df_1[,-1],
             gap = 0,
             bg = c("red", "blue")[df_1$Gender],
             pch=21)

pairs.panels(gender_pca$x,
             gap=0,
             bg = c("red", "blue")[df_1$Gender],
             pch=21)




fviz_eig(gender_pca, addlabels = TRUE)
fviz_pca_var(gender_pca,col.var = "cos2",
             gradient.cols = c("#FFCC00", "#CC9933", "#660033", "#330033"),
             repel = TRUE)
fviz_pca_ind(gender_pca, col.ind = "cos2", 
                  gradient.cols = c("#FFCC00", "#CC9933", "#660033", "#330033"), 
                  repel = TRUE)
biplot(gender_pca)
autoplot(gender_pca,
         data = df_1[,-1],
         loadings = TRUE,
         labels = df_1$Gender)

# Different PCA Method. 
res.pca <- PCA(df_1[,-1], graph = FALSE)
print(res.pca)

# Visualize and Interpret PCA using these functions 

#get_eigenvalue(res.pca): Extract the eigenvalues/variances of principal components
#fviz_eig(res.pca): Visualize the eigenvalues
#get_pca_ind(res.pca), get_pca_var(res.pca): Extract the results for individuals and variables, respectively.
#fviz_pca_ind(res.pca), fviz_pca_var(res.pca): Visualize the results individuals and variables, respectively.
#fviz_pca_biplot(res.pca): Make a biplot of individuals and variables.

eig.val <- get_eigenvalue(res.pca)
eig.val

fviz_eig(res.pca, addlabels = TRUE, ylim = c(0, 50))

var <- get_pca_var(res.pca)
#var$coord: coordinates of variables to create a scatter plot
#var$cos2: represents the quality of representation for variables on the factor map. It’s calculated as the squared coordinates: var.cos2 = var.coord * var.coord.
#var$contrib: contains the contributions (in percentage) of the variables to the principal components. 
#The contribution of a variable (var) to a given principal component is (in percentage) : (var.cos2 * 100) / (total cos2 of the component).
var

# Coordinates
head(var$coord)
# Cos2: quality on the factore map
head(var$cos2)
# Contributions to the principal components
head(var$contrib)

#The plot Below is also known as variable correlation plots. It shows the relationships between all variables. It can be interpreted as follow:

#Positively correlated variables are grouped together.
#Negatively correlated variables are positioned on opposite sides of the plot origin (opposed quadrants).
#The distance between variables and the origin measures the quality of the variables on the factor map. 
#Variables that are away from the origin are well represented on the factor map.

# Correlation circle
fviz_pca_var(res.pca, col.var = "black")

# Quality of representation


corrplot(var$cos2, is.corr=FALSE)
# Total cos2 of variables on Dim.1 and Dim.2
#A high cos2 indicates a good representation of the variable on the principal component. 
#In this case the variable is positioned close to the circumference of the correlation circle.
#A low cos2 indicates that the variable is not perfectly represented by the PCs. 
#In this case the variable is close to the center of the circle.

fviz_cos2(res.pca, choice = "var", axes = 1:2)
fviz_pca_var(res.pca, col.var = "cos2",
             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"), 
             repel = TRUE # Avoid text overlapping
             )
# Change the transparency by cos2 values
fviz_pca_var(res.pca, alpha.var = "cos2")
corrplot(var$contrib, is.corr=FALSE)
# Contributions of variables to PC1
fviz_contrib(res.pca, choice = "var", axes = 1, top = 10)
# Contributions of variables to PC2
fviz_contrib(res.pca, choice = "var", axes = 2, top = 10)
fviz_pca_var(res.pca, col.var = "contrib",
             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07")
             )
fviz_pca_var(res.pca, alpha.var = "contrib")

fviz_pca_ind(res.pca,
             geom.ind = "point", # show points only (nbut not "text")
             col.ind = df_1$Gender, # color by groups
             palette = c("#00AFBB", "#E7B800", "#FC4E07"),
             addEllipses = TRUE, # Concentration ellipses
             legend.title = "Groups"
             )


# Description of PC

res.desc <- dimdesc(res.pca, axes = c(1,2,3,4), proba = 0.05)
# Description of dimension 1
res.desc$Dim.1
res.desc$Dim.2
res.desc$Dim.3
res.desc$Dim.4

# Graph of Indiviuals
ind <- get_pca_ind(res.pca)
ind

## Principal Component Analysis Results for individuals
##  ===================================================
##   Name       Description                       
## 1 "$coord"   "Coordinates for the individuals" 
## 2 "$cos2"    "Cos2 for the individuals"        
## 3 "$contrib" "contributions of the individuals"
#To get access to the different components, use this:

# Coordinates of individuals
head(ind$coord)
# Quality of individuals
head(ind$cos2)
# Contributions of individuals
head(ind$contrib)

fviz_pca_ind(res.pca)

fviz_pca_ind(res.pca, col.ind = "cos2", 
             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"),
             repel = TRUE # Avoid text overlapping (slow if many points)
             )
fviz_pca_ind(res.pca, pointsize = "cos2", 
             pointshape = 21, fill = "#E7B800",
             repel = TRUE # Avoid text overlapping (slow if many points)
             )

fviz_pca_ind(res.pca, col.ind = "cos2", pointsize = "cos2",
             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"),
             repel = TRUE # Avoid text overlapping (slow if many points)
             )
fviz_cos2(res.pca, choice = "ind")
# Total contribution on PC1 and PC2
fviz_contrib(res.pca, choice = "ind", axes = 1:2)

# Create a random continuous variable of length 23,
# Same length as the number of active individuals in the PCA
set.seed(123)
my.cont.var <- rnorm(131)
# Color individuals by the continuous variable
fviz_pca_ind(res.pca, col.ind = my.cont.var,
             gradient.cols = c("blue", "yellow", "red"),
             legend.title = "Cont.Var")

fviz_pca_ind(res.pca,
             geom.ind = "point", # show points only (nbut not "text")
             col.ind = df_1$Gender, # color by groups
             palette = c("#00AFBB", "#E7B800", "#FC4E07"),
             addEllipses = TRUE, # Concentration ellipses
             legend.title = "Groups"
             )

fviz_pca_ind(res.pca, geom.ind = "point", col.ind = df_1$Gender, 
             palette = c("#00AFBB", "#E7B800", "#FC4E07"),
             addEllipses = TRUE, ellipse.type = "confidence",
             legend.title = "Groups"
             )
fviz_pca_ind(res.pca,
             label = "none", # hide individual labels
             habillage = df_1$Gender, # color by groups
             addEllipses = TRUE, # Concentration ellipses
             palette = "jco"
             )
fviz_pca_var(res.pca, geom.var = c("point", "text"))
# Show individuals text labels only
fviz_pca_ind(res.pca, geom.ind =  "text")
# Change the size of arrows an labels
fviz_pca_var(res.pca, arrowsize = 1, labelsize = 5, 
             repel = TRUE)
# Change points size, shape and fill color
# Change labelsize
fviz_pca_ind(res.pca, 
             pointsize = 3, pointshape = 21, fill = "lightblue",
             labelsize = 5, repel = TRUE)

fviz_pca_ind(res.pca,
             geom.ind = "point", # show points only (but not "text")
             group.ind = df_1$Gender, # color by groups
             legend.title = "Groups",
             mean.point = FALSE)
fviz_pca_ind(res.pca,
             geom.ind = "point", # show points only (but not "text")
             group.ind = df_1$Gender, # color by groups
             legend.title = "Groups",
             mean.point = TRUE)
fviz_pca_var(res.pca, axes.linetype = "blank")



ind.p <- fviz_pca_ind(res.pca, geom = "point", col.ind = df_1$Gender)
ggpubr::ggpar(ind.p,
              title = "Principal Component Analysis",
              subtitle = "Gender data set",
              caption = "Source: factoextra",
              xlab = "PC1", ylab = "PC2",
              legend.title = "Gender", legend.position = "top",
              ggtheme = theme_gray(), palette = "jco"
              )

fviz_pca_biplot(res.pca, repel = TRUE,col.ind = df_1$Gender,
                col.var = "#2E9FDF", # Variables color
                )

fviz_pca_biplot(res.pca, 
                col.ind = df_1$Gender, palette = "jco", 
                addEllipses = TRUE, label = "var",
                col.var = "black", repel = TRUE,
                legend.title = "Gender") 

fviz_pca_biplot(res.pca, 
                # Fill individuals by groups
                geom.ind = "point",
                pointshape = 21,
                pointsize = 2.5,
                fill.ind = df_1$Gender,
                col.ind = "black",
                # Color variable by groups
                legend.title = list(fill = "Gender", color = "Clusters"),
                repel = TRUE        # Avoid label overplotting
             )+
  ggpubr::fill_palette("jco")+      # Indiviual fill color
  ggpubr::color_palette("npg")      # Variable colors

fviz_pca_biplot(res.pca, 
                # Individuals
                geom.ind = "point",
                fill.ind = df_1$Gender, col.ind = "black",
                pointshape = 21, pointsize = 2,
                palette = "jco",
                addEllipses = TRUE,
                # Variables
                alpha.var ="contrib", col.var = "contrib",
                gradient.cols = "RdYlBu",
                
                legend.title = list(fill = "Gender", color = "Contrib",
                                    alpha = "Contrib")
                )

## http://www.sthda.com/english/articles/31-principal-component-methods-in-r-practical-guide/112-pca-principal-component-analysis-essentials/

```
